---
title: "Gender Classification"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Gender Classification |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/elakiricoder/gender-classification-dataset/data\>*

### Reference:

*\<ElakiriCoder. (2023). Gender Classification Dataset [Data set]. Kaggle. https://www.kaggle.com/datasets/elakiricoder/gender-classification-dataset\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Understanding the Dataset (Exploratory Data Analysis (EDA))
## Loading the Dataset
```{r Load dataset}
# Load dataset
GenderData <- read.csv("gender_classification_v7.csv", colClasses = c(
  long_hair = "factor",
  forehead_width_cm = "numeric",
  forehead_height_cm = "numeric",
  nose_wide = "factor",
  nose_long = "factor",
  lips_thin = "factor",
  distance_nose_to_lip_long = "factor",
  gender = "factor"
))

# Preview structure
str(GenderData)
head(GenderData)
View(GenderData)
```

## Measures of Frequency
```{r MOF}
#Measures of Frequency
# Frequency of categorical variables
table(GenderData$gender)
table(GenderData$long_hair)
table(GenderData$nose_wide)
table(GenderData$nose_long)
table(GenderData$lips_thin)
table(GenderData$distance_nose_to_lip_long)
```

## Measures of Central Tendency
```{r MOCT}
# Measures of Central Tendency
# Mean and median
mean(GenderData$forehead_width_cm)
median(GenderData$forehead_width_cm)

mean(GenderData$forehead_height_cm)
median(GenderData$forehead_height_cm)

# Mode function (since base R doesn't have one)
get_mode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

get_mode(GenderData$forehead_width_cm)
get_mode(GenderData$forehead_height_cm)
```

## Measures of Distribution
```{r MOD}
# Measures of Distribution
# Distribution metrics
sd(GenderData$forehead_width_cm)
range(GenderData$forehead_width_cm)
IQR(GenderData$forehead_width_cm)

sd(GenderData$forehead_height_cm)
range(GenderData$forehead_height_cm)
IQR(GenderData$forehead_height_cm)
```

## Measures of Relationship
```{r MOR}
# Measures of Relationship
# Correlation between numeric variables
cor(GenderData$forehead_width_cm, GenderData$forehead_height_cm)

# Chi-square test between categorical variables and gender
chisq.test(table(GenderData$long_hair, GenderData$gender))
chisq.test(table(GenderData$nose_wide, GenderData$gender))
chisq.test(table(GenderData$nose_long, GenderData$gender))
chisq.test(table(GenderData$lips_thin, GenderData$gender))
chisq.test(table(GenderData$distance_nose_to_lip_long, GenderData$gender))
```

## ANOVA
```{r ANOVA}
# ANOVA for forehead_width_cm by gender
anova_fw <- aov(forehead_width_cm ~ gender, data = GenderData)
summary(anova_fw)

# ANOVA for forehead_height_cm by gender
anova_fh <- aov(forehead_height_cm ~ gender, data = GenderData)
summary(anova_fh)

```

## Plots
```{r Plots}
#Plots
library(ggplot2)

# Bar plots for categorical variables
ggplot(GenderData, aes(x = long_hair)) + 
  geom_bar(fill = "skyblue") + 
  ggtitle("Frequency of Long Hair")

ggplot(GenderData, aes(x = nose_wide)) + 
  geom_bar(fill = "tomato") + 
  ggtitle("Frequency of Wide Nose")

ggplot(GenderData, aes(x = lips_thin)) + 
  geom_bar(fill = "plum") + 
  ggtitle("Frequency of Thin Lips")

# Histograms for numeric variables
ggplot(GenderData, aes(x = forehead_width_cm)) + 
  geom_histogram(fill = "steelblue", bins = 15) + 
  ggtitle("Distribution of Forehead Width (cm)")

ggplot(GenderData, aes(x = forehead_height_cm)) + 
  geom_histogram(fill = "darkseagreen", bins = 15) + 
  ggtitle("Distribution of Forehead Height (cm)")


# Boxplots to compare numeric variables by gender
ggplot(GenderData, aes(x = gender, y = forehead_width_cm, fill = gender)) + 
  geom_boxplot() + 
  ggtitle("Forehead Width by Gender")

ggplot(GenderData, aes(x = gender, y = forehead_height_cm, fill = gender)) + 
  geom_boxplot() + 
  ggtitle("Forehead Height by Gender")

# Faceted bar plots for categorical features by gender
ggplot(GenderData, aes(x = long_hair, fill = gender)) + 
  geom_bar(position = "dodge") + 
  ggtitle("Long Hair by Gender")

ggplot(GenderData, aes(x = nose_long, fill = gender)) + 
  geom_bar(position = "dodge") + 
  ggtitle("Long Nose by Gender")

```

# Preprocessing & Data Transformation
## Missing Values
```{r Missing Values}
# Check for any missing values
anyNA(GenderData)

# Count of missing values per column
colSums(is.na(GenderData))

# Visual representation (optional)
library(Amelia)  # for missmap
missmap(GenderData, main = "Missing Values Map", col = c("yellow", "black"), legend = FALSE)

# Alternative visual with VIM
# install.packages("VIM") # Uncomment if not installed
library(VIM)
aggr(GenderData, col = c("navyblue", "red"), numbers = TRUE, sortVars = TRUE,
     labels = names(GenderData), cex.axis = .7, gap = 3,
     ylab = c("Missing Data", "Pattern"))
```

# Training Models
## Data Splitting
```{r Data Splitting}
# Load required library
library(caret)

# Set seed for reproducibility
set.seed(123)

# Create a training and testing split (80% train, 20% test)
splitIndex <- createDataPartition(GenderData$gender, p = 0.8, list = FALSE)

# Create training and testing datasets
train_data <- GenderData[splitIndex, ]
test_data  <- GenderData[-splitIndex, ]

# Check the dimensions to confirm
cat("Training Data Size: ", nrow(train_data), "\n")
cat("Testing Data Size: ", nrow(test_data), "\n")
```

## Bootstrapping
```{r Bootstrapping}
# Load required library
library(caret)

# Set seed for reproducibility
set.seed(123)

# Create bootstrap resampling with 1000 resamples
ctrl <- trainControl(method = "boot", number = 1000, savePredictions = "final")

# Train a logistic regression model using bootstrapping
boot_model <- train(gender ~ ., 
                    data = train_data, 
                    method = "glm", 
                    family = "binomial", 
                    trControl = ctrl, 
                    metric = "Accuracy")  # Use Accuracy for evaluation

# Print the model results
print(boot_model)

# Check the resampling results
boot_results <- boot_model$resample
summary(boot_results)
```

## Training Different Models
```{r Training different Models}
# Train a logistic regression model
logistic_model <- train(gender ~ ., 
                        data = train_data, 
                        method = "glm", 
                        family = "binomial", 
                        trControl = ctrl, 
                        metric = "Accuracy")  # Use Accuracy as the evaluation metric

# Print the results of the logistic regression model
print(logistic_model)

# Train a Random Forest model with fewer trees
rf_model <- train(gender ~ ., 
                  data = train_data, 
                  method = "rf", 
                  trControl = ctrl, 
                  metric = "Accuracy", 
                  tuneGrid = expand.grid(mtry = 2),  # you can specify other mtry values here
                  ntree = 100)  # Reduce the number of trees


# Print the results of the random forest model
print(rf_model)
```

## Performance Comparison Using Resamples
```{r resamples}
# Load caret package if not already loaded
library(caret)

# Compare model performance
model_results <- resamples(list(
  Logistic = logistic_model,
  RandomForest = rf_model
))

# Summary of the comparison
summary(model_results)

# Boxplots to visualize accuracy and kappa
bwplot(model_results, metric = "Accuracy")
bwplot(model_results, metric = "Kappa")

# Dotplots as alternative visualization
dotplot(model_results, metric = "Accuracy")

```

## Saving Model
```{r Saving Model}
# Save the best performing model
saveRDS(rf_model, "models/saved_rf_model.rds")

# Load the saved model
loaded_rf_model <- readRDS("models/saved_rf_model.rds")

# Example new data for prediction (update with actual values as needed)
new_data <- data.frame(
  long_hair = factor(1, levels = levels(GenderData$long_hair)),
  forehead_width_cm = 13.5,
  forehead_height_cm = 6.2,
  nose_wide = factor(0, levels = levels(GenderData$nose_wide)),
  nose_long = factor(1, levels = levels(GenderData$nose_long)),
  lips_thin = factor(0, levels = levels(GenderData$lips_thin)),
  distance_nose_to_lip_long = factor(1, levels = levels(GenderData$distance_nose_to_lip_long))
)

# Make prediction
prediction <- predict(loaded_rf_model, newdata = new_data)

# Print prediction
print(prediction)

```

